On the power of conditional independence testing \\ under model-X

### 2022-10-29 arxiv-v5/crtpower-arxiv.tex
Finalized arxiv v5

\begin{abstract}
For testing conditional independence (CI) of a response $Y$ and a predictor $X$ given covariates $Z$, the recently introduced model-X (MX) framework has been the subject of active methodological research, especially in the context of MX knockoffs and their successful application to genome-wide association studies. In this paper, we study the power of MX CI tests, \textcolor{blue}{yielding quantitative insights into the role of machine learning and providing evidence in favor of using likelihood-based statistics in practice. Focusing on the conditional randomization test (CRT), we find that its conditional mode of inference allows us to reformulate it as testing a point null hypothesis involving the conditional distribution of $X$. The Neyman-Pearson lemma then implies that a likelihood-based statistic yields the most powerful CRT against a point alternative.} We also obtain a related optimality result for MX knockoffs. Switching to an asymptotic framework with arbitrarily growing covariate dimension, we derive an expression for the limiting power of the CRT against local semiparametric alternatives in terms of the prediction error of the machine learning algorithm on which its test statistic is based. Finally, we exhibit a resampling-free test with uniform asymptotic Type-I error control under the assumption that \textit{only the first two moments of $X$ given $Z$ are known}, a significant relaxation of the MX assumption. 
\end{abstract}

@@

As formalized by Shah and Peters \cite{Shah2018}, the problem \eqref{conditional-independence} is fundamentally impossible without assumptions on the distribution $\mathcal L(\prx,\pry, \prz)$, in which case no asymptotically uniformly valid test of this hypothesis can have nontrivial power against \emph{any} alternative. In special cases, the problem is more tractable, for example if $\prz$ has discrete support, or if we were willing to make (semi)parametric assumptions on the form of $\mathcal L(\pry|\prx, \prz)$ (henceforth ``model-$Y|X$''). We will not be making such assumptions in this work.
% A complementary recent line of work has focused on shifting assumptions to the distribution $f_{\prx, \prz}$. In particular, 
Instead, we follow the lead of Candes et al. \cite{CetL16}, who proposed to avoid assumptions on $\mathcal L(\pry|\prx, \prz)$, but assume that we have access to $\mathcal L(\prx|\prz)$:\footnote{Candes et al. actually require that the full joint distribution $\mathcal L(\prx,\prz)$ is known, but this is because they also test for conditional associations between $\prz$ and $\pry$. We focus only on the relationship between $\prx$ and $\pry$ given $\prz$ and therefore require a weaker assumption.} 
\begin{equation} \label{eq:modelX}
\textit{model-X (MX) assumption}: \mathcal L(\prx| \prz)=  f^*_{\prx|\prz} \text{ for some known } f^*_{\prx|\prz}.\footnote{\textcolor{blue}{We implicitly assume that $\mathcal L$ has a density with respect to some dominating measure on $\mathbb R^{1+1+p}$, and that all conditional densities are well-defined almost surely. Here and throughout the paper, we identify probability distributions with their densities with respect to the appropriate dominating measure.}} 
\end{equation}
Candes et al.\ argue that while both model-$Y|X$ and MX are strong assumptions---especially when $p,d$ are large---in certain cases much more is known about $\prx | \prz$ than about $\pry|\prx, \prz$. In the aforementioned GWAS example, $\prx|\prz$ reflects the joint distribution of genotypes at SNPs across the genome, which is well described by hidden Markov models from population genetics \cite{SetC17}. On the other hand, the distribution $\pry|\prx,\prz$ represents the genetic basis of a complex trait, about which much less is known. In the context of (stratified) randomized experiments, the distribution $\mathcal L(\prx | \prz)$ is the propensity function \cite{Imai2004} (the analog of the propensity score for non-binary treatments \cite{Rosenbaum1983}) and is experimentally controlled. In general causal inference contexts, the MX assumption can be viewed as the assumption that the propensity function is known.

@@

We start with any test statistic $T(\srx, \sry, \srz)$ measuring the association between $\prx$ and $\pry$, given $\prz$. Usually, this statistic involves learning some estimate $\widehat f_{\pry|\prx,\prz}$ based on machine learning, e.g. the magnitude of the fitted coefficient for $\prx$ (when $\text{dim}(\prx) = 1$) in a cross-validated lasso~\cite{T96} of $\sry$ on $\srx$ and $\srz$ \cite{CetL16}. To calculate the distribution of $T$ under the null hypothesis~\eqref{conditional-independence}, first define a matrix  $\srxk \in \mathbb R^{n \times d}$, where the $i$th row $\srxk_i$ is a sample from $\mathcal L(\prx \mid \prz = \srz_i)$. In other words, for each sample $i$, resample $\srx_i$ based on its distribution conditional on the observed covariate values $Z_i$ in that sample. We then use these resamples to build a null distribution $T(\srxk, \sry, \srz)$, from which the upper quantile
\begin{equation}
\color{blue}C(\sry,\srz)\color{black} \equiv Q_{1-\alpha}[T(\srxk, \sry, \srz)|\sry,\srz]
	\label{upper-quantile}
\end{equation}
may be extracted \textcolor{blue}{(the dependence on $\alpha$ left implicit)}, where the randomness is over the resampling distribution $\srxk |  \sry, \srz$. Finally, the CRT rejects if the original test statistic exceeds this quantile:

@@

While the composite nonparametric alternative of the CI problem~\eqref{conditional-independence} suggests that we cannot expect to find a uniformly most powerful test, we may still ask what is the most powerful test against a point alternative. \textcolor{blue}{Restricting our attention to tests valid conditionally on $(\sry, \srz)$ (as the CRT is)} allows us to reduce the composite null to a point null. We can therefore apply the Neyman-Pearson lemma to show (Section~\ref{sec:power}) that the optimal conditionally valid test against a point alternative $\mathcal L$ with $\mathcal L(\pry|\prx,\prz) = \bar f_{\pry|\prx,\prz}$ is the CRT based on the likelihood test statistic:
\begin{equation}
	T^{\text{opt}}(\srx; \sry, \srz) \equiv \prod_{i = 1}^n \bar f(\sry_i|\srx_i, \srz_i).
\end{equation}
The same statistic yields the most powerful one-bit $p$-values for MX knockoffs (Section~\ref{sec:knockoffs}). Despite the simplicity of this result, it has not been derived before and appears central to the design of powerful test statistics. Since the model for $Y|X,Z$ is unknown, this result provides our first theoretical indication of the usefulness of machine learning models to learn this distribution (Q2). A2 below gives a more quantitative answer to Q2.

@@

In this section, we seek the most powerful CRT against a point alternative. To accomplish this, we make the observation---\textcolor{blue}{implicit in earlier works}---that the CRT is valid not just unconditionally but also conditionally on $\sry, \srz$ (Section~\ref{sec:bridge}). The latter conditioning step reduces the composite null to a point null. This reduction allows us to invoke the Neyman Pearson lemma to find the most powerful test  (Section~\ref{sec:NP}). Proofs are deferred to the appendix.

@@

Let us first formalize the definition of a level $\alpha$ test of the MX CI problem. The null hypothesis is defined as the set of joint distributions compatible with conditional independence and with the assumed model for $\prx|\prz$:
\begin{equation}
	\label{eq:null-under-modelX}
	\begin{split}
		\mathscr L^{\text{MX}}_0(f^*) &\equiv \mathscr L_0 \cap \mathscr L^{\text{MX}}(f^*) \\
		&\equiv \{\mathcal L: \prx \independent \pry \mid \prz\} \cap \{\mathcal L: \mathcal L(\prx|\prz) = f^*_{\prx|\prz}\}\\
		&= \{\mathcal L: \mathcal L(\prx,\pry,\prz) = f_{\prz} \cdot f^*_{\prx|\prz} \cdot f_{\pry|\prz} \text{ for some } f_{\prz}, f_{\pry|\prz} \}.
	\end{split}
\end{equation}
A test $\phi: (\mathbb R^{d} \times \mathbb R^r \times \mathbb R^p)^n \rightarrow [0,1]$ of the MX CI problem is said to be level $\alpha$ if
\begin{equation}
	\sup_{\mathcal L \in \mathscr L^{\text{MX}}_0(f^*)} \mathbb E_{\mathcal L}[\phi(\srx, \sry, \srz)] \leq \alpha.
	\label{marginal-validity}
\end{equation}
Recall that the CRT critical value $C(\sry, \srz)$ is defined via conditional calibration~\eqref{upper-quantile}. As is known to those familiar with MX, this implies that any CRT $\phi = \phi_T^{\text{CRT}}$ not only has level $\alpha$ in the sense of definition~\eqref{marginal-validity} but also has level $\alpha$ \textit{conditionally} on $\sry$ and $\srz$:
\color{blue}
\begin{equation}
\sup_{\mathcal L \in \mathscr L_0^{\textnormal{MX}}(f^*)}\ \mathbb E_{\mathcal L}[\phi(\srx, \sry, \srz)|\sry, \srz] \leq \alpha \quad \text{almost surely}.
\label{eq:conditional-validity}
\end{equation}
\color{black}

One special property of such conditionally valid tests $\phi$ is that they can be viewed as testing a \textit{point null} rather than the original \textit{composite null} \eqref{conditional-independence}. To see this, we view $\phi \equiv \phi(\srx;\sry,\srz)$ as a \textit{family} of hypothesis tests, indexed by $(\sry, \srz)$, for the distribution $\mathcal L(\srx|\sry,\srz)$. Note that under the MX assumption,
\small
\begin{equation}
\mathcal L \in \mathscr L_0^{\text{MX}}(f^*) \Longrightarrow \mathcal L(\srx = \sfx|\sry = \sfy, \srz = \sfz) = \prod_{i = 1}^n  f^*(\sfx_i | \sfz_i).
\label{point-null}
\end{equation}
\normalsize
In words, fixing $Y,Z$ at their realizations $\sfy,\sfz$ and viewing only $\srx$ as random, $\mathcal L(\srx|\sry = \sfy, \srz = \sfz)$ equals a fixed product distribution for any null $\mathcal L$. This yields a conditional point null hypothesis, with respect to which $\phi^{\CRT}_T(\sfx;\sfy,\sfz)$ is a level-$\alpha$ test for \textcolor{blue}{almost every} $(\sfy, \sfz)$. Note that the observations $\srx_i$ in this conditional distribution are independent \textit{but not identically distributed} due to the different conditioning events in~\eqref{point-null}.

\color{blue}
We emphasize that the aforementioned observations have been under the hood of MX papers, and the existence of a single null distribution from which to resample $\srxk$ is central to the very definition of the CRT. Nevertheless, we find it useful to state explicitly what has thus far been largely left implicit. Indeed, viewing the CRT through the conditional lens~\eqref{eq:conditional-validity} is the starting point that allows us to bring classical theoretical tools to bear on its analysis. We start doing so by considering point alternatives below.
\color{black}

@@

%The form of the optimal statistic $T^{\text{opt}}$ may be surprising, as it is a likelihood ratio with respect to $\sry|\srx,\srz$ rather than $\srx|\sry,\srz$. In fact, we see in the proof (Appendix~\ref{sec:proofs-sec2}) that these two likelihood ratios are equivalent, with the former yielding a more explicit expression. Note also that the denominator of $T^{\text{opt}}$ can be omitted if desired, since it is not a function of $\srx$. We keep it for a more direct parallel with the model-$\sry|\srx$ case. % Though this paper is primarily concerned with the CRT, we also obtain a parallel result about MX knockoffs in Section~\ref{sec:knockoffs}. 
\textcolor{blue}{We leave open the question of whether $\phi^{\CRT}_{T^\textnormal{opt}}$ is also the most powerful test among not just conditionally valid tests~\eqref{eq:conditional-validity} but also among marginally valid tests~\eqref{marginal-validity}. There do at least exist marginally valid tests that are not conditionally valid.}

@@

Theorem~\ref{prop:crt-optimality} and Corollary~\ref{cor:crt-optimality} imply that the most powerful CRT against a point alternative is based on the test statistic defined as the measuring how well the data $(\srx, \sry, \srz)$ fit the distribution $\bar{\mathcal L}(\pry|\prx,\prz)$. For example, if
\begin{equation}
\bar f(\pry|\prx,\prz) = N(\prx^T\beta + \prz^T \gamma, \sigma^2) \text{ for coefficients } \beta \in \mathbb R^d \text{ and } \gamma \in \mathbb R^p,
\label{linear-model}
\end{equation}
then the optimal test rejects for small values of $\|\sry - \srx \beta - \srz \gamma \|^2$. In Section~\ref{sec:knockoffs}, we establish an analogous optimality statement for MX knockoffs as well. Since the optimal test depends on the alternative distribution $\bar{\mathcal L}(\pry|\prx,\prz)$, CRT and MX knockoffs implementations usually employ a machine learning step to search through the composite alternative (not unlike a likelihood ratio test) for a good approximation $\widehat f_{\pry|\prx,\prz}$. These approximate models are then summarized in various ways to define a test statistic $T$. There is no consensus yet on the best test statistic to use, with some authors \cite{CetL16, SetC17, SetS19} using combinations of fitted coefficients $\widehat \beta$ and others  \cite{Tansey2018, Bates2020} using likelihood-based test statistics. The above optimality results align more closely with the latter strategy. \textcolor{blue}{Theorem~\ref{prop:crt-optimality} has inspired an extension of the CRT to the sequential setting using a likelihood-based test statistic, accompanied by a similar optimality result~\cite{Grunwald2022}.} Likelihood-based test statistics also have the advantage of avoiding ad hoc combination rules for $\widehat \beta \in \mathbb R^d$ when $d > 1$. It remains to be seen whether likelihood-based or coefficient-based test statistics yield greater power in practice, but a thorough empirical comparison is beyond the scope of this work. For now, it suffices to note that, despite its simplicity, this is the first such power optimality result in the CRT literature.

@@

Now, consider a set of semiparametric \textcolor{blue}{(partially linear)} alternatives $\mathcal L_n(\pry|\prx,\prz)$ such that
\begin{equation}
	\pry = \prx^T \beta_n + g_n(\prz) + \peps; \quad \peps \sim N(0, \sigma^2 ),\ \sigma^2 > 0
	\label{eq:linearity}
\end{equation}
for $\peps \independent (\prx, \pry, \prz)$. Here, $\beta_n \in \mathbb R^d$ is a coefficient vector, $g_n: \mathbb R^p \rightarrow \mathbb R$ a general function, and $\sigma^2 > 0$ the residual variance. Of special interest are local alternatives where $\beta_n = h/\sqrt{n}$ for some $h \in \mathbb R^d$. \textcolor{blue}{We emphasize that---in this section and throughout the paper---we use the partially linear model~\eqref{eq:linearity} exclusively as an alternative distribution against which to assess power, rather than an additional assumption required for Type-I error control.} By Theorem~\ref{prop:crt-optimality}, the most powerful test against the alternative~\eqref{eq:linearity} is the CRT based on the likelihood statistic 
\begin{equation}
\begin{split}
T_n^{\text{opt}}(\srx, \sry, \srz) &= \prod_{i = 1}^n\frac{1}{(2\pi)^{1/2}}\exp\left(-\frac{1}{2\sigma^2}\left(\sry_i - \srx_i^T \beta_n - g_n(\srz_i)\right)^2\right) \\
&=\prod_{i = 1}^n\frac{1}{(2\pi)^{1/2}}\exp\left(-\frac{1}{2\sigma^2}\left(\sry_i - (\srx_i-\mu_n(\srz_i))^T \beta_n - g_n'(\srz_i)\right)^2\right),
\label{eq:likelihood-ratio}
\end{split}
\end{equation}
where
\begin{equation}
\color{blue}\bar g_n(\prz)\color{black} \equiv \mathbb E[\pry|\prz] =  \mu_n(\prz)^T\beta_n  + g_n(\prz).
\label{eq:g-n-prime-def}
\end{equation}
Assuming local alternatives $\beta_n = h/\sqrt n$ and taking a logarithm, we obtain
\begin{equation}
	\begin{split}
		\log T_n^{\text{opt}}(\srx, \sry, \srz) &= -\frac n 2 \log(2\pi) - \frac{1}{2\sigma^2}\sum_{i = 1}^n\left(\sry_i - (\srx_i-\mu_n(\srz_i))^T h/\sqrt{n} - \bar g_n(\srz_i)\right)^2 \\
		&\approx   \frac{h^T}{\sigma^2}\frac{1}{\sqrt{n}}\sum_{i = 1}^n(\sry_i - \bar g_n(\srz_i))(\srx_i-\mu_n(\srz_i)) + C,
	\end{split}
\label{eq:optimal-semiparametric}
\end{equation}
where $C$ is a constant that does not depend on $\srx$ and therefore does not change upon resampling.

@@

Informally, this theorem (proved in Appendix~\ref{sec:proofs-sec45}) suggests that the CRT resampling distribution of $T_n(\srx, \sry, \srz)$ converges to $\chi^2_d$, which is the resampling distribution of this test statistic under a normal $\mathcal L_n(\prx|\prz)$. Note that the resulting equivalence~\eqref{eq:asymptotic-equivalence} holds for the specific instance of the CRT based on the statistic $T_n$ defined in via equations~\eqref{u-hat} and~\eqref{t-n}, though other kinds of test statistics may lead to similar large-sample behavior. While Theorem~\ref{thm:equivalence} is stated for $\widehat g_n$ trained out of sample, we conjecture that it continues to hold when $\widehat g_n$ is fit in sample, as in the original dCRT construction \cite{Liu2020}. \textcolor{blue}{At least, we observe that the conditioning in the construction of the resampling distribution $\mathcal L_n(\sqrt n \cdot \widehat \rho_n(\srxk, \sry, \srz) \mid \srx, \sry, \srz)$ ensures that its mean and variance remain equal to $0$ and $\widehat S_n^2$ even when $\widehat g_n$ is fit in sample.}

@@

See Appendix~\ref{sec:proofs-sec45} for a proof of this theorem. The moment assumptions can be relaxed for pointwise error control (less desirable), but are unavoidable for uniform type-I error control as stated in the corollary. More importantly, we conjecture that the MX(2) $F$-test continues to have asymptotic Type-I error control even if $\widehat g_n$ is fit in sample. \textcolor{blue}{One may expect this because the validity of the MX(2) $F$-test derives from the correctness of $(\mu_n, \Sigma_n)$ rather than that of $\widehat g_n$.} This conjecture is supported by the results of a simulation study \textcolor{blue}{presented in Appendix~\ref{sec:simulations}}.

@@

\textcolor{blue}{Having found that the dCRT is a natural test to apply for power against semiparametric alternatives, and that this test is equivalent to the simpler MX(2) $F$-test, we are ready to study the relationship between the power of the dCRT and the quality of the underlying machine learning procedure.}

>>>>

\begin{abstract}
For testing conditional independence (CI) of a response $Y$ and a predictor $X$ given covariates $Z$, the recently introduced model-X (MX) framework has been the subject of active methodological research, especially in the context of MX knockoffs and their successful application to genome-wide association studies. In this paper, we study the power of MX CI tests, yielding quantitative insights into the role of machine learning and providing evidence in favor of using likelihood-based statistics in practice. Focusing on the conditional randomization test (CRT), we find that its conditional mode of inference allows us to reformulate it as testing a point null hypothesis involving the conditional distribution of $X$. The Neyman-Pearson lemma then implies that a likelihood-based statistic yields the most powerful CRT against a point alternative. We also obtain a related optimality result for MX knockoffs. Switching to an asymptotic framework with arbitrarily growing covariate dimension, we derive an expression for the limiting power of the CRT against local semiparametric alternatives in terms of the prediction error of the machine learning algorithm on which its test statistic is based. Finally, we exhibit a resampling-free test with uniform asymptotic Type-I error control under the assumption that \textit{only the first two moments of $X$ given $Z$ are known}, a significant relaxation of the MX assumption. 
\end{abstract}

@@

As formalized by Shah and Peters \cite{Shah2018}, the problem \eqref{conditional-independence} is fundamentally impossible without assumptions on the distribution $\mathcal L(\prx,\pry, \prz)$, in which case no asymptotically uniformly valid test of this hypothesis can have nontrivial power against \emph{any} alternative. In special cases, the problem is more tractable, for example if $\prz$ has discrete support, or if we were willing to make (semi)parametric assumptions on the form of $\mathcal L(\pry|\prx, \prz)$ (henceforth ``model-$Y|X$''). We will not be making such assumptions in this work.
% A complementary recent line of work has focused on shifting assumptions to the distribution $f_{\prx, \prz}$. In particular, 
Instead, we follow the lead of Candes et al. \cite{CetL16}, who proposed to avoid assumptions on $\mathcal L(\pry|\prx, \prz)$, but assume that we have access to $\mathcal L(\prx|\prz)$:\footnote{Candes et al. actually require that the full joint distribution $\mathcal L(\prx,\prz)$ is known, but this is because they also test for conditional associations between $\prz$ and $\pry$. We focus only on the relationship between $\prx$ and $\pry$ given $\prz$ and therefore require a weaker assumption.} 
\begin{equation} \label{eq:modelX}
\textit{model-X (MX) assumption}: \mathcal L(\prx| \prz)=  f^*_{\prx|\prz} \text{ for some known } f^*_{\prx|\prz}.\footnote{We implicitly assume that $\mathcal L$ has a density with respect to some dominating measure on $\mathbb R^{1+1+p}$, and that all conditional densities are well-defined almost surely. Here and throughout the paper, we identify probability distributions with their densities with respect to the appropriate dominating measure.}
\end{equation}
Candes et al.\ argue that while both model-$Y|X$ and MX are strong assumptions---especially when $p,d$ are large---in certain cases much more is known about $\prx | \prz$ than about $\pry|\prx, \prz$. In the aforementioned GWAS example, $\prx|\prz$ reflects the joint distribution of genotypes at SNPs across the genome, which is well described by hidden Markov models from population genetics \cite{SetC17}. On the other hand, the distribution $\pry|\prx,\prz$ represents the genetic basis of a complex trait, about which much less is known. In the context of (stratified) randomized experiments, the distribution $\mathcal L(\prx | \prz)$ is the propensity function \cite{Imai2004} (the analog of the propensity score for non-binary treatments \cite{Rosenbaum1983}) and is experimentally controlled. In general causal inference contexts, the MX assumption can be viewed as the assumption that the propensity function is known.

@@

We start with any test statistic $T(\srx, \sry, \srz)$ measuring the association between $\prx$ and $\pry$, given $\prz$. Usually, this statistic involves learning some estimate $\widehat f_{\pry|\prx,\prz}$ based on machine learning, e.g. the magnitude of the fitted coefficient for $\prx$ (when $\text{dim}(\prx) = 1$) in a cross-validated lasso~\cite{T96} of $\sry$ on $\srx$ and $\srz$ \cite{CetL16}. To calculate the distribution of $T$ under the null hypothesis~\eqref{conditional-independence}, first define a matrix  $\srxk \in \mathbb R^{n \times d}$, where the $i$th row $\srxk_i$ is a sample from $\mathcal L(\prx \mid \prz = \srz_i)$. In other words, for each sample $i$, resample $\srx_i$ based on its distribution conditional on the observed covariate values $Z_i$ in that sample. We then use these resamples to build a null distribution $T(\srxk, \sry, \srz)$, from which the upper quantile
\begin{equation}
C(\sry,\srz) \equiv Q_{1-\alpha}[T(\srxk, \sry, \srz)|\sry,\srz]
	\label{upper-quantile}
\end{equation}
may be extracted (the dependence on $\alpha$ left implicit), where the randomness is over the resampling distribution $\srxk |  \sry, \srz$. Finally, the CRT rejects if the original test statistic exceeds this quantile:

@@

While the composite nonparametric alternative of the CI problem~\eqref{conditional-independence} suggests that we cannot expect to find a uniformly most powerful test, we may still ask what is the most powerful test against a point alternative. Restricting our attention to tests valid conditionally on $(\sry, \srz)$ (as the CRT is) allows us to reduce the composite null to a point null. We can therefore apply the Neyman-Pearson lemma to show (Section~\ref{sec:power}) that the optimal conditionally valid test against a point alternative $\mathcal L$ with $\mathcal L(\pry|\prx,\prz) = \bar f_{\pry|\prx,\prz}$ is the CRT based on the likelihood test statistic:
\begin{equation}
	T^{\text{opt}}(\srx; \sry, \srz) \equiv \prod_{i = 1}^n \bar f(\sry_i|\srx_i, \srz_i).
\end{equation}
The same statistic yields the most powerful one-bit $p$-values for MX knockoffs (Section~\ref{sec:knockoffs}). Despite the simplicity of this result, it has not been derived before and appears central to the design of powerful test statistics. Since the model for $Y|X,Z$ is unknown, this result provides our first theoretical indication of the usefulness of machine learning models to learn this distribution (Q2). A2 below gives a more quantitative answer to Q2.

@@

In this section, we seek the most powerful CRT against a point alternative. To accomplish this, we make the observation---implicit in earlier works---that the CRT is valid not just unconditionally but also conditionally on $\sry, \srz$ (Section~\ref{sec:bridge}). The latter conditioning step reduces the composite null to a point null. This reduction allows us to invoke the Neyman Pearson lemma to find the most powerful test  (Section~\ref{sec:NP}). Proofs are deferred to the appendix.

@@

Let us first formalize the definition of a level $\alpha$ test of the MX CI problem. The null hypothesis is defined as the set of joint distributions compatible with conditional independence and with the assumed model for $\prx|\prz$:
\begin{equation}
	\label{eq:null-under-modelX}
	\begin{split}
		\mathscr L^{\text{MX}}_0(f^*) &\equiv \mathscr L_0 \cap \mathscr L^{\text{MX}}(f^*) \\
		&\equiv \{\mathcal L: \prx \independent \pry \mid \prz\} \cap \{\mathcal L: \mathcal L(\prx|\prz) = f^*_{\prx|\prz}\}\\
		&= \{\mathcal L: \mathcal L(\prx,\pry,\prz) = f_{\prz} \cdot f^*_{\prx|\prz} \cdot f_{\pry|\prz} \text{ for some } f_{\prz}, f_{\pry|\prz} \}.
	\end{split}
\end{equation}
A test $\phi: (\mathbb R^{d} \times \mathbb R^r \times \mathbb R^p)^n \rightarrow [0,1]$ of the MX CI problem is said to be level $\alpha$ if
\begin{equation}
	\sup_{\mathcal L \in \mathscr L^{\text{MX}}_0(f^*)} \mathbb E_{\mathcal L}[\phi(\srx, \sry, \srz)] \leq \alpha.
	\label{marginal-validity}
\end{equation}
Recall that the CRT critical value $C(\sry, \srz)$ is defined via conditional calibration~\eqref{upper-quantile}. As is known to those familiar with MX, this implies that any CRT $\phi = \phi_T^{\text{CRT}}$ not only has level $\alpha$ in the sense of definition~\eqref{marginal-validity} but also has level $\alpha$ \textit{conditionally} on $\sry$ and $\srz$:
\begin{equation}
\sup_{\mathcal L \in \mathscr L_0^{\textnormal{MX}}(f^*)}\ \mathbb E_{\mathcal L}[\phi(\srx, \sry, \srz)|\sry, \srz] \leq \alpha \quad \text{almost surely}.
\label{eq:conditional-validity}
\end{equation}

One special property of such conditionally valid tests $\phi$ is that they can be viewed as testing a \textit{point null} rather than the original \textit{composite null} \eqref{conditional-independence}. To see this, we view $\phi \equiv \phi(\srx;\sry,\srz)$ as a \textit{family} of hypothesis tests, indexed by $(\sry, \srz)$, for the distribution $\mathcal L(\srx|\sry,\srz)$. Note that under the MX assumption,
\small
\begin{equation}
\mathcal L \in \mathscr L_0^{\text{MX}}(f^*) \Longrightarrow \mathcal L(\srx = \sfx|\sry = \sfy, \srz = \sfz) = \prod_{i = 1}^n  f^*(\sfx_i | \sfz_i).
\label{point-null}
\end{equation}
\normalsize
In words, fixing $Y,Z$ at their realizations $\sfy,\sfz$ and viewing only $\srx$ as random, $\mathcal L(\srx|\sry = \sfy, \srz = \sfz)$ equals a fixed product distribution for any null $\mathcal L$. This yields a conditional point null hypothesis, with respect to which $\phi^{\CRT}_T(\sfx;\sfy,\sfz)$ is a level-$\alpha$ test for almost every $(\sfy, \sfz)$. Note that the observations $\srx_i$ in this conditional distribution are independent \textit{but not identically distributed} due to the different conditioning events in~\eqref{point-null}.

We emphasize that the aforementioned observations have been under the hood of MX papers, and the existence of a single null distribution from which to resample $\srxk$ is central to the very definition of the CRT. Nevertheless, we find it useful to state explicitly what has thus far been largely left implicit. Indeed, viewing the CRT through the conditional lens~\eqref{eq:conditional-validity} is the starting point that allows us to bring classical theoretical tools to bear on its analysis. We start doing so by considering point alternatives below.

@@

%The form of the optimal statistic $T^{\text{opt}}$ may be surprising, as it is a likelihood ratio with respect to $\sry|\srx,\srz$ rather than $\srx|\sry,\srz$. In fact, we see in the proof (Appendix~\ref{sec:proofs-sec2}) that these two likelihood ratios are equivalent, with the former yielding a more explicit expression. Note also that the denominator of $T^{\text{opt}}$ can be omitted if desired, since it is not a function of $\srx$. We keep it for a more direct parallel with the model-$\sry|\srx$ case. % Though this paper is primarily concerned with the CRT, we also obtain a parallel result about MX knockoffs in Section~\ref{sec:knockoffs}. 
We leave open the question of whether $\phi^{\CRT}_{T^\textnormal{opt}}$ is also the most powerful test among not just conditionally valid tests~\eqref{eq:conditional-validity} but also among marginally valid tests~\eqref{marginal-validity}. There do at least exist marginally valid tests that are not conditionally valid.

@@

Theorem~\ref{prop:crt-optimality} and Corollary~\ref{cor:crt-optimality} imply that the most powerful CRT against a point alternative is based on the test statistic defined as the measuring how well the data $(\srx, \sry, \srz)$ fit the distribution $\bar{\mathcal L}(\pry|\prx,\prz)$. For example, if
\begin{equation}
\bar f(\pry|\prx,\prz) = N(\prx^T\beta + \prz^T \gamma, \sigma^2) \text{ for coefficients } \beta \in \mathbb R^d \text{ and } \gamma \in \mathbb R^p,
\label{linear-model}
\end{equation}
then the optimal test rejects for small values of $\|\sry - \srx \beta - \srz \gamma \|^2$. In Section~\ref{sec:knockoffs}, we establish an analogous optimality statement for MX knockoffs as well. Since the optimal test depends on the alternative distribution $\bar{\mathcal L}(\pry|\prx,\prz)$, CRT and MX knockoffs implementations usually employ a machine learning step to search through the composite alternative (not unlike a likelihood ratio test) for a good approximation $\widehat f_{\pry|\prx,\prz}$. These approximate models are then summarized in various ways to define a test statistic $T$. There is no consensus yet on the best test statistic to use, with some authors \cite{CetL16, SetC17, SetS19} using combinations of fitted coefficients $\widehat \beta$ and others  \cite{Tansey2018, Bates2020} using likelihood-based test statistics. The above optimality results align more closely with the latter strategy. Theorem~\ref{prop:crt-optimality} has inspired an extension of the CRT to the sequential setting using a likelihood-based test statistic, accompanied by a similar optimality result~\cite{Grunwald2022}. Likelihood-based test statistics also have the advantage of avoiding ad hoc combination rules for $\widehat \beta \in \mathbb R^d$ when $d > 1$. It remains to be seen whether likelihood-based or coefficient-based test statistics yield greater power in practice, but a thorough empirical comparison is beyond the scope of this work. For now, it suffices to note that, despite its simplicity, this is the first such power optimality result in the CRT literature.

@@

Now, consider a set of semiparametric (partially linear) alternatives $\mathcal L_n(\pry|\prx,\prz)$ such that
\begin{equation}
	\pry = \prx^T \beta_n + g_n(\prz) + \peps; \quad \peps \sim N(0, \sigma^2 ),\ \sigma^2 > 0
	\label{eq:linearity}
\end{equation}
for $\peps \independent (\prx, \pry, \prz)$. Here, $\beta_n \in \mathbb R^d$ is a coefficient vector, $g_n: \mathbb R^p \rightarrow \mathbb R$ a general function, and $\sigma^2 > 0$ the residual variance. Of special interest are local alternatives where $\beta_n = h/\sqrt{n}$ for some $h \in \mathbb R^d$. We emphasize that---in this section and throughout the paper---we use the partially linear model~\eqref{eq:linearity} exclusively as an alternative distribution against which to assess power, rather than an additional assumption required for Type-I error control. By Theorem~\ref{prop:crt-optimality}, the most powerful test against the alternative~\eqref{eq:linearity} is the CRT based on the likelihood statistic 
\begin{equation}
\begin{split}
T_n^{\text{opt}}(\srx, \sry, \srz) &= \prod_{i = 1}^n\frac{1}{(2\pi)^{1/2}}\exp\left(-\frac{1}{2\sigma^2}\left(\sry_i - \srx_i^T \beta_n - g_n(\srz_i)\right)^2\right) \\
&=\prod_{i = 1}^n\frac{1}{(2\pi)^{1/2}}\exp\left(-\frac{1}{2\sigma^2}\left(\sry_i - (\srx_i-\mu_n(\srz_i))^T \beta_n - g_n'(\srz_i)\right)^2\right),
\label{eq:likelihood-ratio}
\end{split}
\end{equation}
where
\begin{equation}
\bar g_n(\prz) \equiv \mathbb E[\pry|\prz] =  \mu_n(\prz)^T\beta_n  + g_n(\prz).
\label{eq:g-n-prime-def}
\end{equation}
Assuming local alternatives $\beta_n = h/\sqrt n$ and taking a logarithm, we obtain
\begin{equation}
	\begin{split}
		\log T_n^{\text{opt}}(\srx, \sry, \srz) &= -\frac n 2 \log(2\pi) - \frac{1}{2\sigma^2}\sum_{i = 1}^n\left(\sry_i - (\srx_i-\mu_n(\srz_i))^T h/\sqrt{n} - \bar g_n(\srz_i)\right)^2 \\
		&\approx   \frac{h^T}{\sigma^2}\frac{1}{\sqrt{n}}\sum_{i = 1}^n(\sry_i - \bar g_n(\srz_i))(\srx_i-\mu_n(\srz_i)) + C,
	\end{split}
\label{eq:optimal-semiparametric}
\end{equation}
where $C$ is a constant that does not depend on $\srx$ and therefore does not change upon resampling.

@@

Informally, this theorem (proved in Appendix~\ref{sec:proofs-sec45}) suggests that the CRT resampling distribution of $T_n(\srx, \sry, \srz)$ converges to $\chi^2_d$, which is the resampling distribution of this test statistic under a normal $\mathcal L_n(\prx|\prz)$. Note that the resulting equivalence~\eqref{eq:asymptotic-equivalence} holds for the specific instance of the CRT based on the statistic $T_n$ defined in via equations~\eqref{u-hat} and~\eqref{t-n}, though other kinds of test statistics may lead to similar large-sample behavior. While Theorem~\ref{thm:equivalence} is stated for $\widehat g_n$ trained out of sample, we conjecture that it continues to hold when $\widehat g_n$ is fit in sample, as in the original dCRT construction \cite{Liu2020}. At least, we observe that the conditioning in the construction of the resampling distribution $\mathcal L_n(\sqrt n \cdot \widehat \rho_n(\srxk, \sry, \srz) \mid \srx, \sry, \srz)$ ensures that its mean and variance remain equal to $0$ and $\widehat S_n^2$ even when $\widehat g_n$ is fit in sample.

@@

See Appendix~\ref{sec:proofs-sec45} for a proof of this theorem. The moment assumptions can be relaxed for pointwise error control (less desirable), but are unavoidable for uniform type-I error control as stated in the corollary. More importantly, we conjecture that the MX(2) $F$-test continues to have asymptotic Type-I error control even if $\widehat g_n$ is fit in sample. One may expect this because the validity of the MX(2) $F$-test derives from the correctness of $(\mu_n, \Sigma_n)$ rather than that of $\widehat g_n$. This conjecture is supported by the results of a simulation study presented in Appendix~\ref{sec:simulations}.

@@

Having found that the dCRT is a natural test to apply for power against semiparametric alternatives, and that this test is equivalent to the simpler MX(2) $F$-test, we are ready to study the relationship between the power of the dCRT and the quality of the underlying machine learning procedure.
